Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-20 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small touch-ups and re-formulations in summary
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2941 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 19def0c8742f8f7e869c5227c4f31e0ee1102b0e

The performance of the novel circular regression trees and forests is assessed in a case
study for shortterm probabilistic wind direction forecasting at two airports
with different site characteristics. As benchmark models, probabilistic
climatology and persistency models, as well as a state-of-the-art circular
GLM-type model are evaluated based on proper scoring rules. In summary, the
circular regression trees and forests have the highest predictive performance in this
setting. For cases without changes in the wind regime, lagged response values
provide already highly skillful estimates leading to a good performance of the
persistency model as observed for shortterm wind direction forecasts in this
study. While in these cases the trees and forests also benefit from the highly
informative lagged response, they can compensate for a lower information of
this covariate by incorporating other quantities and possible interactions of
these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods. For the specific task of
wind forecasting not only the wind direction, but also the wind speed and
especially their combination is of high interest. To account for both
quantities simultaneously, distributional tree-based models could be built for
wind vectors employing the bivariate normal distribution as performed by
\citet{Lang+Mayr+Stauffer:2019} in an additive regression framework.
Alternatively, circular regression trees could be combined with distributional models for
wind speed employing the truncated normal distribution
\citep{Thorarinsdottir+Gneiting:2010}. This could be pursued either in a
two-step approach with two independent distributional models or in a more
sophisticated way employing a mixed response distribution in the presented
tree-building algorithm. Another possible improvement for obtaining more
parsimonious circular regression trees is to consider splitting circular covariates into
two circle segments by searching two split points simultaneously rather than
sequentially at different depths. While this might slightly improve the
predictive performance of circular regression trees, this should not affect the
performance of the forests, as they consist of very large trees with many
different splits.

>>>>

The performance of the novel circular regression trees and forests is assessed in a case
study for shortterm probabilistic wind direction forecasting at two airports
with different site characteristics. As benchmark models, probabilistic
climatology and persistency models, as well as a state-of-the-art circular
GLM-type model are evaluated based on proper scoring rules. In summary, the
circular regression trees and forests have the highest predictive performance in this
setting. For cases without changes in the wind regime, lagged response values
provide already highly skillful estimates leading to a good performance of the
persistency model as observed for shortterm wind direction forecasts in this
study. While in these cases the trees and forests also benefit from the highly
informative lagged response, they can compensate for a lower information of
this covariate by incorporating other quantities and possible interactions of
these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods.

For the specific task of wind forecasting, the wind direction is often only
relevant if the wind speed is sufficiently high. Hence, it is of interest
to account for both quantities simultaneously, e.g., by considering a 
bivariate normal distribution for the wind vectors (from which wind speed
and wind direction can be obtained). The parameters of this bivariate normal
distribution could then be linked to the available covariate using 
an additive regression framework \citep[as proposed by][]{Lang+Mayr+Stauffer:2019}
or using a tree-based approach, similar to the one proposed in this paper.
Moreover, a rather different approach for a combined response of wind speed
and wind direction would be a two-step or hurdle model: In the first step
this could build on the truncated normal model of
\cite{Thorarinsdottir+Gneiting:2010} to capture wind speed. And in a second
step a circular wind direction model is leveraged provided that a certain
hurdle for the wind speed is crossed.

Another possible improvement for obtaining more
parsimonious circular regression trees is to consider splitting circular covariates into
two circle segments by searching two split points simultaneously rather than
sequentially at different depths. While this might slightly improve the
predictive performance of circular regression trees, this should not affect the
performance of the forests, as they consist of very large trees with many
different splits.
